Title here

### 2020-06-08 paper/paper.Rmd
add DFO in introduction; rewrite introduction + vis diagnostics

## Tour

Tour provides a way to explore multivariate data interactively via an established tour path. A tour path is formed by interpolating between randomly generated plane. Different types of tours are available depends on the purpose of the exploration i.e. a grand tour is suitable for randomly exploring the data from different angles; a guided tour detects a particular structure in the data. a manual tour allows user to manually control the projection [@cook1997manual].

## Guided tour

Guided tour is usually used in conjunction projection pursuit, a method coined by @friedman1974projection to detect "interesting" low-dimension projection of multivariate data. A projection pursuit requires the definition of a projection pursuit index function and an optimisation routine. The projection pursuit index measures the "interestingness" of data defined as its departure from normality. Numerous indices have been proposed in the literature, including lengendre index [@friedman1974projection], hermite index [@hall1989polynomial], natural hermite index [@cook1993projection], chi-square index [@posse1995projection], LDA index [@lee2005projection] and PDA index [@lee2010projection]. An optimisation routine is required to find the projection basis (thus projection) that maximises the projection pursuit index. The discussion of existing optimisation procedure will be discussed in the next section.

Guided tour creates visualisation for the projections found by projection pursuit by constructing a tour path. An illustration modified from [@buja2005computational] entails the geodesic interpolation between the plane generated by projection pursuit.

```{r tour-path}
knitr::include_graphics(path = here::here("figures/tour_path_new.png"))
```

## Optimisation methods in projection pursuit literature

As @friedman1974projection said "..., the technique use for maximizing the (one- and two-dimensional) projection index strongly influences both the statistical and the computational aspects of the procedure." The quality of the optimisation procedure largely affect the tour view and thus, the interesting projection one could possibly observe. An ideal optimisation procedure needs to have following characteristics:

- *Being able to handle non-differentiable index function*: the index function could be noisy and non-differentiable

- *Being able to optimise with constraints*: the projection matrix is restricted to an orthonormal matrix.

- *Being able to reveal both local and global maximum*: Although the primary interest is to find the global maximum, distinct local structures are also of our interest.

Below present three existing methods in tour.

@posse1995projection presented a random search algorithm that samples new basis in the neighbourhood of the current basis. The neighbourhood is defined via the radius of the p-dimensional sphere, $c$. The new basis is taken as the target basis if it has higher index value, or the sampling continues. If no basis is found to have higher index value after a certain number of tries $n$, the radius $c$ is halved. The algorithm stops when the maximum number of iteration is attained or the radius $c$ is less than a pre-determined number. [Pursuit package uses this method and it works great! But I don't think we implement this - although don't think it is too hard to do it].

@@

<!-- Because of the fact that projection pursuit index can be non-differentiable and that the projection basis is constrained to be orthonormal, the optimisation need to be a derivative-free, constrained procedure. Also, local maxima is equally interested as global maxima since it reveals distinct local structure in the data. -->

@@

# Iterative algorithm and its diagnostics

## Tour components

Guided tour, along with other types of tour, has been implemented in the *tourr* package in R, available on the Comprehensive R Archive Network at [https://cran.r-project.org/web/packages/tourr/](https://cran.r-project.org/web/packages/tourr/) [@wickham2011tourrpackage]. A tour includes two major components: a *generator* that generating the projection basis according to projection pursuit and an *interpolator* that performing geodesic interpolation between the projection basis.

The pseudo-code below illustrates the implementation of guided tour in the tourr package. Given an projection pursuit index function and a randomly generated projection basis (current basis), the optimisation procedure produces a target basis inside `generator()`  Both the current basis and the target basis will be supplied to `tour_path()` to prepare information needed for constructing a geodesic path. This information is then used to compute a series of interpolating bases inside the `tour()` function. All the basis will be sent to create animation for visualising the tour in the `animate()` function.

```{r algo,eval = FALSE, echo = TRUE}
animation <- function(){

  # compute projection basis 
  tour <- function(){

    # construct bases on the tour path
    new_geodesic_path <- function(){
      tour_path <- function(){

        # GENERATOR: generate projection basis via projection pursuit
        guided_tour <- function(){
          generator <- function(){

            # define projection pursuit index
            # generate the target basis from the current basis via optimisation
          }
        }

        # prepare geodesic information needed for interpolating along the tour path
      }
    }

    # INTERPOLATOR: interpolate between the current and target basis 
    function(){
      # generate interpolating bases on the geodesic path
    }
  }

  # animate according to different display methods
}
```

## Diagnostics

Visualisation has been widely used for exploring and understanding data. Visualisation presents information in a graphical manner and often allows people to see information they would otherwise not seen in the reporting of numerical summarisation. Visual diagnostics can be real-time or post-run. Real time diagnostic directly uses the data produced in the algorithm to produce visual representation and thus doesn't need to store the data.   This section focuses on the definition and production of post-run diagnostics and the next section discusses real-time diagnostics.

Post-run diagnostics requires the data produced during the algorithm to be stored/ saved in order to produce plot diagnostics.

The graphical system in R is established based on the conception grammar of graphic [@wickham2010layered], where a graphic is defined using stacked layers in a coordination system.  A layer includes 1) the dataset that powers the plot; 2) a geometric object that represents the visual shape of the plot and 3) relevant statistical transformation if needed. An important concept in the grammar of graphic is *aesthetic mapping*. Aesthetic mapping links the variable in a dataset to information needed to produce a geometric object. For example, we map one variable on the x-axis and another on the y-axis to create a scatterplot. To create a boxplot, we first map one variable on the x-axis and then map the five point summary of another variable on the y-axis. This computation of five point summary from the origin variable is statistical transformation. This definition of graphic through layers provides advantages to produce complex plots since all the plots, however complex, can be decomposed into basic geometric objects shaped by the variable supplied.

This idea of decompositing a graph to basic elements inspires me to characterise the diagnostics of iterative algorithm in a similar fashion. The optimisation in many machine learning algorithms these days is iterative in nature, while remains as a black box. Being able to diagnose it visually allows researchers to have the tool to unfold the myth and thus provides more understand to the algorithm.

The concept of grammar of graphic requires a dataset to be supplied in making a plot, thus a global object needs to be created in the iterative algorithm. The variables will be mapped into element of a graphic to explore and thus should contain all the parameters of interest. The next section shows how a global object is created in projection pursuit guided tour.

## Data Structure

In the current implementation of the `tourr` package, while the target basis generated by the projection pursuit can be accessed later via `save_history()`, interpolating bases and those randomly nearby bases generated in the optimisation are not stored. This creates difficulties for fully understand the behaviour of the optimisation and interpolation of tour in complex scenario **[need a rephrase this part].**

Two set of simulated data are used in the demonstration of the visualisation and diagnostics of the tour optimisation. A small dataset consists of 1000 randomly simulated observations of five variables (`x1`, `x2`, `x8`, `x9`, `x10`). `x2` is the informative variable simulated from two bi-modal normal distribution centred at -3 and 3 with variance being 1 and the other four are simulated from `N(0, 1)`. The data has been scaled to ensure `x2` has variance of 1.

A larger dataset contains more informative variables (`x3` to `x7`) of different types. `x3` takes 500 positive one and 500 negative one. The distribution of all the variables except `x3` is plotted below. *[should I introduce the dist for each var?]*

@@

# Post-run static diagnostics plots

Given all the information of interest in a global object, the next thing is to design its visual representation that can diagnose our algorithm. In an iterative algorithm, one would be interested to know how one parameter changes as the algorithm iterates. The parameter could be could be the decision variable, the value of the optimisation function or other parameter that facilitate the algorithm. In projection pursuit guided tour, apart from the decision variable projection basis, the value of index function (`index_val`), we are also interested to explore the effect the different searching methods, the neighbourhood parameter alpha, and the stage a particular observation is in (`info`). These are the variables that will be mapped into the diagnostic plot. The global object also has its time series nature since the observations are recorded as the algorithm progresses and thus can't be switch in rows. This time series structure is always hierarchical and nested since each row can be labelled with the smallest unit `id` and up by the iterative structure of the algorithm. Take our example, in the global object, each row has an `id` label which is the smallest unit the observation is ordered. A larger ordering unit is `tries` since it is updated every time a new target basis is found. A nested ordering unit is `loop`, which increases by one as searching method iterates and starts over at a new `tries`. These features explain the definition of the global object for projection pursuit guided tour in Figure \ref{fig:glb-obj}.

```{r glb-obj, out.width= "100%", out.height="20%", fig.cap="\\label{glb-obj}this is xxxx"}
knitr::include_graphics(path = here::here("figures/global_obj.png"))
```

@@

*Example: exploring searching points* The data is sourced from a 2D projection of the larger dataset and `search_better` is used with `max.tries = 500`. In Figure \ref{fig:points-id} and \ref{fig:points-tries}, a comparison has been made on visualising the searching points with `id` and `tries` on the x-axis, colored by `tries`. In Figure \ref{fig:points-id}, the searching points of the first few `tries` are squeezed in a small width leaving the large uninteresting searching points in the last three `tries` taking a vast majority of the plotting space. While in Figure \ref{fig:points-tries} the data is spaced by `tries` evenly in the plot. Label at the bottom indicates the number of observations in each tries and facilitates the choice of cutoff to switch from point geometry to boxplot geometry (`cutoff = 15`). The line geometry suggest the largest improvement happens at `tries = 5`.

@@

Matrix parameter could also be interested to explore, for example, in the tour, we are interested to explore the relative position of the projection basis in the vector space. A projection basis is a vector/matrix whose row number is the number of variable in the given dataset. This imposes difficulties in visualisation since we are bounded to perceive at most three dimensions. Thus, principal component analysis is used to reduce the dimension of the projection basis and the first two principal component are mapped to the x and y axis of the plot. Another variable of interest could be mapped to the color aesthetics to see how it changes as the projection basis changes in the projected 2D space. [have a second read here]

>>>>

## Derivative free optimisation

Given an objective function $f$, one way of optimising it is to equate its gradient to zero. In modern optimisation problems, gradient information can be hard to evaluate or sometimes even impossible and Derivative-Free Optimisation (DFO) methods can be useful to approach these problems. The two most common methods in DFO are *direct-search methods* and *model-based methods* and this paper dedicates to the discussion of direct-search methods, which gains its popularity due to its simplicity in use and reliability in complicated practical problem. Coined by @hooke1961direct, direct search methods don't require any gradient or hessian information and evaluate $f$ directly. A well-known example of it is the Nelder-Mead algorithm [@nelder1965simplex] and it enjoys the popularity due to its simplicity and reliability. [feel like this sentence can be expanded to include more information. xxx]. Further, direct search methods can be classified as *stochastic* and *deterministic* depends on whether a random sample is used in the search. The stochastic version of  direct-search method is the focus of this paper.

## Projection pursuit guided tour

The optimisation problem we're interested in is in the context of projection pursuit. Coined by @friedman1974projection, projection pursuit is a method that detects the interesting structure (i.e. clustering, outliers and skewness) of multivariate data via projecting it in lower dimensions. A Projection Pursuit Index (PPI) is a function of the projection matrix (a.k.a projection angle, projection basis) measuring the "interestingness" of data and we refer the value of PPI function as *index value*. In the literature, the indices being proposed include lengendre index [@friedman1974projection], hermite index [@hall1989polynomial], natural hermite index [@cook1993projection], chi-square index [@posse1995projection], LDA index [@lee2005projection] and PDA index [@lee2010projection]. [Any literature on holes index? xxx]

 As @friedman1974projection noted "..., the technique use for maximising the projection index strongly influences both the statistical and the computational aspects of the procedure." A suitable optimisation procedure is needed to find the projection angle that maximises the PPI and the quality of the optimisation largely affect the interesting projections one could possibly observe.

Projection pursuit is usually used in conjunction with a tour method called *guided tour*. Tour explores the multivariate data *interactively* via playing a series of projections, that form a *tour path*, like movie frames in real time.  Starting from a randomly generated projection basis, projection pursuit searches for a target basis with higher index value and guided tour interpolates geodesically between the initial basis and the target basis.  This series of bases forms a tour path and the projection of the data on these bases is played in real time.  This search-and-interpolateprocess is repeated until no basis can be found with higher index value. Modified from @buja2005computational, Figure \ref{tour-path} vividly depicts the tour path in guided tour. Guided tour, along with other types of tour, has been implemented in the *tourr* package in R, available on the Comprehensive R Archive Network at [https://cran.r-project.org/web/packages/tourr/](https://cran.r-project.org/web/packages/tourr/) [@wickham2011tourrpackage].

```{r tour-path, out.width="100%", out.height="60%", fig.cap="\\label{tour-path}An illustration of the tour path"}
knitr::include_graphics(path = here::here("figures/tour_path_increasing.png"))
```

[doubt if I need the following. xxx]
<!-- A tour path includes two major components: a *generator* that generating the target basis that has higher index value than the current basis and an *interpolator* that performing geodesic interpolation between the current and target basis. An illustration modified from [@buja2005computational] vividly depicts the tour path in guided tour. -->

<!-- The pseudo-code below illustrates the implementation of guided tour in the tourr package. Given an projection pursuit index function and a randomly generated projection basis (current basis), the optimisation procedure produces a target basis inside `generator()`  Both the current basis and the target basis will be supplied to `tour_path()` to prepare information needed for constructing a geodesic path. This information is then used to compute a series of interpolating bases inside the `tour()` function. All the basis will be sent to create animation for visualising the tour in the `animate()` function.   -->

<!-- ```{r algo,eval = FALSE, echo = TRUE} -->
<!-- animation <- function(){ -->

<!--   # compute projection basis  -->
<!--   tour <- function(){ -->

<!--     # construct bases on the tour path -->
<!--     new_geodesic_path <- function(){ -->
<!--       tour_path <- function(){ -->

<!--         # GENERATOR: generate projection basis via projection pursuit -->
<!--         guided_tour <- function(){ -->
<!--           generator <- function(){ -->

<!--             # define projection pursuit index -->
<!--             # generate the target basis from the current basis via optimisation -->
<!--           } -->
<!--         } -->

<!--         # prepare geodesic information needed for interpolating along the tour path -->
<!--       } -->
<!--     } -->

<!--     # INTERPOLATOR: interpolate between the current and target basis  -->
<!--     function(){ -->
<!--       # generate interpolating bases on the geodesic path -->
<!--     } -->
<!--   } -->

<!--   # animate according to different display methods -->
<!-- } -->
<!-- ``` -->

## Optimisation in projection pursuit

Below listed several features characterise the optimisation procedures needed in projection pursuit

- *Being able to handle non-differentiable PPI function*: The PPI function could be non-differentiable, thus derivative free methods are preferred.

- *Being able to optimise with constraints*: The constraint comes from projection matrix being an orthonormal matrix.

- *Being able to find both local and global maximum*: Although the primary interest is to find the global maximum, local maximum could also reeal structures in the data that are of our interest.

[thinking if the following three should be presented as algorithms or plain description is fine. xxx]

There are three existing methods for optimisating PPI function and we review them below. @posse1995projection used a stochastic direct search method,  a random search algorithm to sample new bases in the neighbourhood of the current basis defined by the radius of the p-dimensional sphere, $c$. The new basis is taken as the target basis if it has higher index value, or the sampling continues. If no basis is found to have higher index value after a certain number of tries $n$, the radius $c$ is halved. The algorithm stops when the maximum number of iteration is attained or the radius $c$ is less than a pre-determined number.

@@

@@

# Visual diagnostic system

## Visual diagnostics 
<!-- Visual diagnostics can be real-time or post-run. Real time diagnostic directly uses the data produced in the algorithm to produce visual representation and thus doesn't need to store the data.   This section focuses on the definition and production of post-run diagnostics and the next section discusses real-time diagnostics. -->

<!-- Visualisation has been widely used for exploring and understanding data. Presenting information in a graphical manner often allows people to see information they would otherwise not seen. -->

Random search methods has a black-box machinism and focuses solely on finding the global maximum point, while the projection pursuit problem we have is aiming at *exploring* the data and thus is interested in how the algorithm finds its maximum. This motivates us **to develop a visual diagnostic system for exploring the optimisation searching path**.

The necessity of developing such a system rather than simply producing different diagnostic plots is because there is always needs to develop a function specific for diagnosing one particular variables and these functions developed are not reusable at all. Thus we want to establish a set of rules that prescribe the diagnostic of iterative algorithms systematically.

The idea of unifying all the diagnostic plots under one framework is inspired by the concept of *gramma of graphic*[@wickham2010layered], which powers the primary graphical system in R, ggplot2 [@ggplot2]. In gramma of graphic, plots are not defined by its appearance (i.e. boxplot, histogram, scatter plot etc) but by "stacked layers". By this design, ggplot doesn't need to develop a gazillion of functions that each produces a different type of plot, instead, it defines layers and allow users to specify the plot through these defined layers.

[Do I need this paragraph to details the gramma of graphic? xxx]
<!-- A layer includes 1) the data that powers the plot; 2) a geometric object that represents the visual shape of the data and 3) relevant statistical transformation that transform the information in the data to the information used to draw the geometric object. An important concept in the grammar of graphic is *aesthetic mapping*. Aesthetic mapping links the variable in a dataset to information needed to produce a geometric object. For example, we map one variable on the x-axis and another on the y-axis to create a scatterplot. Another example of creating a boxplot involves 1) mapping one variable on the x-axis and 2) transforming another variable to its five-point-summary and mapping the five-point-summary to the y-axis.  -->

## Glocal object

Ggplot prescribes how to make a graphic but it doesn't prescribe how to run a diagnostic. [need to have a paragraph or two here to describe how I come up with the design of four components in the global object - need to related to the component in optimisation. xxx].

<!-- The concept of grammar of graphic requires a dataset to be supplied in making a plot, thus a global object needs to be created in the iterative algorithm. The variables will be mapped into element of a graphic to explore and thus should contain all the parameters of interest. The next section shows how a global object is created in projection pursuit guided tour. -->

<!-- In the current implementation of the `tourr` package, while the target basis generated by the projection pursuit can be accessed later via `save_history()`, interpolating bases and those randomly nearby bases generated in the optimisation are not stored. This creates difficulties for fully understand the behaviour of the optimisation and interpolation of tour in complex scenario **[need a rephrase this part].**  -->

A global object used for diagnosing an optimisation is defined with four components: decision variable, value of objective function, iterators and other parameter of interest. In the projection pursuit guided tour, the decision variable is the projection matrix, the value of objective function is index value. There are three iterators: `id` is the smallest ordering unit that increases by one for each observatoin; `tries` is a larger order unit, which is updated once a search-and-interpolate step is finished. A third iterator `loop` is used to record the number of repitition in the search step and starts over from one at the beginning of a new `tries`. Since there are different methods used in the projection pursuit guided tour, we are interested to compare between different methods, thus `method` is created as another parameter of interest. The neighbourhood parameter alpha is always interested in the direct search methods since it controls the size of the sampling space, thus `alpha` is another parameter of interest. `info` is created to label different stage in the searching process and it is the third parameter of interest. A sketch of the global object for projection pursuit guided tour is presented in Figure \ref{fig:glb-obj}.

```{r glb-obj, out.width= "100%", out.height="20%", fig.cap="\\label{glb-obj}The global object in projection pursuit guided tour."}
knitr::include_graphics(path = here::here("figures/global_obj.png"))
```

# Diagnostics plots

## Simulated data

Two set of simulated data are used in the demo of the visual diagnostics in projection pursuit guided tour. A small dataset consists of 1000 randomly simulated observations of five variables (`x1`, `x2`, `x8`, `x9`, `x10`). `x2` is the informative variable simulated from two bi-modal normal distribution centred at -3 and 3 with variance of one and the other four are simulated from non-informative standard random normal distributions. The data has been scaled to ensure `x2` has variance of 1. The goal is to find the 1D projection of bi-modal shape, which correspondes to the projection matrix (vector) of `[0, 1, 0, 0, 0]`.

A larger dataset contains more informative variables (`x3` to `x7`) of different types. The distribution of all the variables except `x3` is plotted below and `x3` takes 500 positive one and 500 negative one. [should I introduce the dist for each var? xxx] [also feel like we don't really use `x3` to `x6`, should we not mention about these? xxx]

@@

@@

*Example: exploring searching points* The data is sourced from a 2D projection of the larger dataset and `search_better` is used with `max.tries = 500`. In Figure \ref{fig:points-id} and \ref{fig:points}, a comparison has been made on visualising the searching points with `id` and `tries` on the x-axis, colored by `tries`. In Figure \ref{fig:points-id}, the searching points of the first few `tries` are squeezed in a small width leaving the large uninteresting searching points in the last three `tries` taking a vast majority of the plotting space. While in Figure \ref{fig:points} the data is spaced by `tries` evenly in the plot. Label at the bottom indicates the number of observations in each tries and facilitates the choice of cutoff to switch from point geometry to boxplot geometry (`cutoff = 15`). The line geometry suggest the largest improvement happens at `tries = 5`.

@@

Sometimes, a parameter to explore could be a vector or matrix instead of scalar, for example, in tour, we are interested to explore the projection basis in the vector space.  This imposes difficulties in visualisation since we are bounded to perceive at most three dimensions. Thus, principal component analysis is used to reduce the dimension of projection bases and the first two principal components are mapped to the x and y axis of the plot. Additional variable of interest could be mapped to the color aesthetics to for exploration.
